---
title: "Counter-factual: no Alert level 4 on Jan 4, 2022"
author: "Steve Walker and Amy Hurford"
date: "12/04/2022"
output: 
  html_document: 
    keep_md: yes
---

Below I try to do a counter-factual simulation assuming no Alert level 4 on January 4, 2022

```{r, message = FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(McMasterPandemic)
library(ggplot2)
library(lubridate)
library(patchwork)
setwd("/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan")
source('get_data.R')
source('functions.R')
fit = readRDS('initial_model/fit.rds')
fitted_data = readRDS('initial_model/fitted_data.rds')
observed_data = readRDS('initial_model/observed_data.rds')

params = read_params("PHAC.csv")
params["N"] = 522453  # nfld population
params = fix_pars(params)

# This gives me errors but then I run the line individually and it's fine
params$beta0 = as.numeric(unname(coef(fit, 'fitted')$params[1]))
params$mu = as.numeric(unname(coef(fit, 'fitted')$params[2]))
params$phi1 = as.numeric(unname(coef(fit, 'fitted')$params[3]))

end_date = max(observed_data$date)+14
#Changing from 90 to 15 improves fit substantially
start_date_offset = 15
start_date = min(observed_data$date) - start_date_offset

# The breakpoint on Jan 4 is removed
params_timevar = data.frame(
  Date = c("2022-03-14"), # dates of breakpoints
  Symbol = "beta0",                     # parameters to vary
  Value = coef(fit, 'fitted')$time_params[2],                    # NA means calibrate to data
  Type = "abs"                          # abs = change to value in Value col
)


state <- make_state(params=params)
res1 <- run_sim(params=params,params_timevar=params_timevar,state=state, start_date=start_date, end_date=end_date)
scenario = data.frame(date=res1$date,H=res1$H, D = res1$death, report = res1$report, cumRep = res1$cumRep)
```
Getting the data ready for plotting
```{r}
hospitalizations = filter(observed_data,var=="H")
fitted_hosp = filter(fitted_data,var=="H")

deaths = filter(observed_data,var=="death")%>%group_by(week = cut(date, "week")) %>% summarise(value = sum(value))
fitted_deaths = filter(fitted_data,var=="death")
cum_deaths = filter(observed_data,var=="death")
cum_deaths = data.frame(date = cum_deaths$date, value=cumsum(cum_deaths$value))

cases = filter(observed_data,var=="report")
fitted_cases = filter(fitted_data,var=="report")
cum_cases = filter(fitted_data,var=="cumRep")
```

```{r,include=FALSE}
gH = ggplot()+
  geom_ribbon(data=fitted_hosp, aes(x=as.Date(date),ymin=lwr, ymax=upr), alpha = 0.5, fill = "aquamarine")+
  geom_point(data=hospitalizations, aes(x=as.Date(date),y=value), col = "aquamarine")+
  geom_line(data=fitted_hosp,aes(x=as.Date(date), y=value), col = "aquamarine")+
  geom_line(data=scenario,aes(x=as.Date(date), y=H), col = "black")+
  ylab("Hospital occupancy")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="aquamarine", lty=2)



gD = ggplot()+
  geom_ribbon(data=fitted_deaths, aes(x=as.Date(date),ymin=7*lwr, ymax=7*upr), alpha = 0.5, fill = "plum3")+
  geom_point(data=deaths, aes(x=as.Date(week),y=value), col = "plum3")+
  geom_line(data=fitted_deaths,aes(x=as.Date(date), y=7*value), col = "plum3")+
  geom_line(data=scenario,aes(x=as.Date(date), y=7*D), col = "black")+
  ylab("deaths (weekly)")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)



gcases = ggplot()+
  geom_ribbon(data=fitted_cases, aes(x=as.Date(date),ymin=lwr, ymax=upr), alpha = 0.5, fill = "tan1")+
  geom_point(data=cases, aes(x=as.Date(date),y=value), col = "tan1")+
  geom_line(data=fitted_cases,aes(x=as.Date(date), y=value), col = "tan1")+
  geom_line(data=scenario,aes(x=as.Date(date), y=report), col = "black")+
  ylab("reported new cases")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)



g_cumRep = ggplot()+
  geom_point(data=cases, aes(x=as.Date(date),y=cumsum(value)), col = "tan1")+
  geom_line(data=cum_cases,aes(x=as.Date(date), y=value), col = "tan1")+
  geom_line(data=scenario,aes(x=as.Date(date), y=cumRep), col = "black")+
  ylab("cumulative cases")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)



g_cumD = ggplot()+
  geom_point(data=cum_deaths, aes(x=as.Date(date),y=value), col = "plum3")+
  geom_line(data=fitted_deaths[!is.na(fitted_deaths$value),],aes(x=as.Date(date), y=cumsum(value)), col = "plum3")+
  geom_line(data=scenario[!is.na(scenario$D),],aes(x=as.Date(date), y=cumsum(D)), col = "black")+
  ylab("cumulative deaths")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)
```

```{r, echo=F, warning=FALSE, fig.height=3}
gcases
gH
gD
```

**Figure 1. The effect of the alert level 4 was to create a two-peaked wave 5 of prolonged duration in Newfoundland and Labrador.** Without the implementation of the Alert level 4 on January 4, 2022 (black lines), Newfoundland and Labrador would have experienced a wave 5 similar to other provinces: with a higher peak and dropping to prevalence in early March. With the implementation of the Alert level 4, and the subsequent easing of restrictions, i.e. phased re-opening beginning on February 14, 2022, Newfoundland and Labrador created a two-peaked wave of longer duration.


```{r,echo=F, warning=FALSE, fig.height=3}
g_cumD + g_cumRep
```

**Figure 2. Despite creating a two-peaked wave, the final size for cases and deaths is only slightly reduced relative to the one-peaked wave had no Alert level 4 been implemented.** Steve are the `cumRep` values correct? Very different from observed data, but the fits are substantially overestimating cases and deaths too and this will get magnified for cumulative values.
