---
title: "Newfoundland and Labrador's two-peaked BA.1 wave"
author: "Amy Hurford and Steve Walker"
date: "12/04/2022"
output: 
  html_document: 
    keep_md: yes
---

```{r, message = FALSE, warning=FALSE, include=FALSE}
library(tidyr)
library(dplyr)
library(McMasterPandemic)
library(ggplot2)
library(lubridate)
library(patchwork)
devtools::source_gist("98cc4db25867bd18cc42b6568b4c6848", sha1 = "3cc333562e")
setwd("/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan")
source('functions.R')
observed_data = readRDS('observed_data.rds')
observed_data2 = readRDS('observed_data2.rds')
fit = readRDS('fit.rds')
fitted_data = readRDS('fitted_data.rds')

params = read_params("PHAC.csv")
params["N"] = 522453  # nfld population
params = fix_pars(params)

# This gives me errors but then I run the line individually and it's fine
params$beta0 = as.numeric(unname(coef(fit, 'fitted')$params[1]))
params$mu = as.numeric(unname(coef(fit, 'fitted')$params[2]))
params$phi1 = as.numeric(unname(coef(fit, 'fitted')$params[3]))

end_date = max(observed_data$date)
#Changing from 90 to 15 improves fit substantially
start_date_offset = 15
start_date = min(observed_data$date) - start_date_offset

# The breakpoint on Jan 4 is removed
params_timevar = data.frame(
  Date = tail(fit$forecast_args$time_args$params_timevar$Date, -1), # dates of breakpoints
  Symbol = "beta0",                     # parameters to vary
  Value = tail(coef(fit, 'fitted')$time_params, -1),                    # NA means calibrate to data
  Type = "abs"                          # abs = change to value in Value col
)


state <- make_state(params=params)
res1 <- run_sim(params=params,params_timevar=params_timevar,state=state, start_date=start_date, end_date=end_date)
scenario = data.frame(date=res1$date,H=res1$H, D = res1$death, report = res1$report, cumRep = res1$cumRep)
```

```{r, include=FALSE}
# Comparison with Ontario
dat=read.csv('https://raw.githubusercontent.com/ahurford/covid-nl/master/covid19-download.csv', fill=TRUE)
datON = dat[dat$prname=="Ontario",]
datON = datON[datON$date>"2021-12-01",]
```

```{r,include=FALSE}
gON = ggplot()+
  geom_ribbon(data=datON,aes(x=as.Date(date), ymax=100*numtoday/max(numtoday), ymin=0), fill = "dodgerblue", alpha = 0.3)+
  geom_line(data=scenario[!is.na(scenario$report),],aes(x=as.Date(date), y=100*report/max(report)), col = "black",lty=1)+
  annotate("text", x = as.Date("2022-03-31"), y = 25, label = "Ontario", col="dodgerblue")+
  annotate("text", x = as.Date("2022-02-10"), y = 75, label = "NL without\nrestrictions")+
  ylab("New cases as % of peak")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")

gH = plot_forecast(fitted_data, "H", observed_data)+ylab("Hospital occupancy")+
  geom_line(data=scenario,aes(x=as.Date(date), y=H), col = "black",lty=1)



deaths = filter(observed_data,var=="death")%>%group_by(week = cut(date, "week")) %>% summarise(value = sum(value))
fitted_deaths = filter(fitted_data,var=="death")

gD = ggplot()+
  geom_ribbon(data=fitted_deaths, aes(x=as.Date(date),ymin=7*lwr, ymax=7*upr), alpha = 0.3, fill = "dodgerblue")+
  geom_point(data=deaths, aes(x=as.Date(week),y=value), fill = "white", pch=21)+
  geom_line(data=fitted_deaths,aes(x=as.Date(date), y=7*value), col = "dodgerblue")+ylab("deaths (weekly)")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)+geom_line(data=scenario,aes(x=as.Date(date), y=7*D), col = "black", lty=1)

observed_cases = filter(observed_data2,var=="report")
gcases = plot_forecast(fitted_data, "report", observed_data)+ylab("Reported cases")+geom_point(data=observed_cases, aes(x=as.Date(date),y=value), col = "dodgerblue", alpha=0.3)+geom_line(data=scenario,aes(x=as.Date(date), y=report), col = "black",lty=1)



```

```{r, echo=F, warning=FALSE, fig.height=2.2}
gcases
gH
gD
```

**Figure 1. The effect of restrictions in early-January in Newfoundland and Labrador was a two-peaked BA.1 Omicron variant wave of prolonged duration that protected health care capacity.** The implementation of the Alert level 4 on January 4, 2022, and the subsequent easing of restrictions on February 14, 2022, meant that Newfoundland and Labrador has experienced a two-peaked BA.1 wave of longer duration. These restrictions helped to protect health care capacity by reducing the number of concurrent hospitalizations. With the restrictions, the total number of cases and deaths was only slightly reduced, suggesting that the predominant effect of the restrictions was to delay, rather than prevent, cases and severe illness.

Below are some awkwardly coded comparisons of cases and deaths.

```{r}
fitted_total_deaths = filter(fitted_data,var=="death")%>%filter(date>"2021-12-15")%>%filter(!is.na(value))%>%summarize(sum(value))

scenario_total_deaths = filter(scenario,date>"2021-12-15")%>%filter(!is.na(D))%>%summarize(sum(D))

fitted_total_cases = filter(fitted_data,var=="report")%>%filter(date>"2021-12-15")%>%filter(!is.na(value))%>%summarize(sum(value))

scenario_total_cases = filter(scenario,date>"2021-12-15")%>%filter(!is.na(report))%>%summarize(sum(report))
```

```{r,echo=F, warning=FALSE, fig.height=2.2}
gON
```

**Figure 2. The BA.1 Omicron variant wave would have been similar to Ontario's had restrictions not been implemented in early January in Newfoundland and Labrador.** The simulated BA.1 Omicron wave for Newfoundland and Labrador without restrictions in early January has a similar shape to Ontario's BA.1 Omicron variant wave, with the wave beginning approximate 3 weeks later.