---
title: "Modelling COVID-19 in Newfoundland and Labrador"
subtitle: Short-term forecasting
author: "Amy Hurford and Steve Walker"
date: "18/04/2022"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This fits a [compartmental epidemic model](https://mcmasterpandemic.shinyapps.io/mcmasterpandemicshiny/) to Newfoundland and Labrador COVID-19 data. This approach is used to model COVID-19 in [Ontario](https://mac-theobio.github.io/forecasts/outputs/McMasterOntarioForecastsBlog2022-01-26), however, there are differences in the model formulation for Newfoundland and Labrador. The software we used can be downloaded [here](https://github.com/mac-theobio/McMasterPandemic).

For an explanation of the assumptions around changes in the transmission rate owing to public health measures, see [here](https://rpubs.com/ahurford/891932).

Further details will be added soon.


```{r, message = FALSE, warning=FALSE, include=FALSE}
library(tidyr)
library(dplyr)
library(McMasterPandemic)
library(ggplot2)
library(lubridate)
library(patchwork)
setwd("/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan")
source('get_data.R')
source('functions.R')
fit = readRDS('/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan/variant_model/variant_model_files/fit_var.rds')
fitted_data = readRDS('/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan/variant_model/variant_model_files/fitted_data_var.rds')
observed_data = readRDS('/Users/ahurford/Desktop/Work/Research/Research_Projects/2022/nfld-macpan/initial_model/initial_model_files/observed_data2.rds')
```



```{r, include=FALSE}
cases = filter(observed_data,var=="report")
fitted_cases = filter(fitted_data,var=="report")
p = var.frac(fitted_cases$date)$var.frac
output = data.frame(date = fitted_cases$date, cases=fitted_cases$value, BA.2 = p*fitted_cases$value)

g.var = ggplot()+
  #geom_point(data=cases,aes(x= as.Date(date),y=value), col = "black", pch=21, fill = "white")+
geom_vline(aes(xintercept = tail(observed_data$date,1)),
                colour = "grey60", linetype = "dashed")+
geom_ribbon(data= output, aes(x= as.Date(date), ymax = cases,ymin=0),fill = "dodgerblue", alpha = 0.3)+
  geom_line(data= output, aes(x= as.Date(date), y = cases),col = "dodgerblue", alpha = 0.3)+
geom_ribbon(data= output, aes(x= as.Date(date), ymax = BA.2,ymin=0),fill = "blue", alpha = 0.3)+
  annotate("text", x = as.Date("2022-03-08"), y = 200, label = "BA.1", col = "dodgerblue")+
    annotate("text", x = as.Date("2022-03-28"), y = 100, label = "BA.2", col="darkblue")+
  ylab("Est. reported new cases (daily)")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b", limits = c(as.Date("2021-12-15"), tail(as.Date(fitted_cases$date),1)))+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")

```

```{r, include=FALSE}
observed_cases = filter(observed_data,var=="report")
gC=plot_forecast(fitted_data, "report", observed_data)+ylab("Est. reported new cases (daily)")+scale_x_date(date_breaks = "7 day", date_labels = "%d %b", limits = c(as.Date("2021-12-15"), tail(as.Date(fitted_cases$date),1)))
#+geom_point(data=observed_cases, aes(x=as.Date(date),y=value), col = "dodgerblue", alpha=0.3)
```


```{r, include=FALSE}
gH=plot_forecast(fitted_data, "H", observed_data)+ylab("Fitted hospital occupancy")+scale_x_date(date_breaks = "7 day", date_labels = "%d %b", limits = c(as.Date("2021-12-15"), tail(as.Date(fitted_cases$date),1)))

```


```{r, include=FALSE}
library(lubridate)
deaths = filter(observed_data,var=="death")%>%group_by(week = cut(date, "week")) %>% summarise(value = sum(value))
fitted_deaths = filter(fitted_data,var=="death")

g1 = ggplot()+
  geom_ribbon(data=fitted_deaths, aes(x=as.Date(date),ymin=7*lwr, ymax=7*upr), alpha = 0.3, fill = "dodgerblue")+
  geom_line(data=fitted_deaths,aes(x=as.Date(date), y=7*value), col = "dodgerblue")+ylab("Fitted deaths (weekly)")+ scale_x_date(date_breaks = "7 day", date_labels = "%d %b", limits = c(as.Date("2021-12-15"), tail(as.Date(fitted_cases$date),1)))+
  geom_point(data=deaths, aes(x=as.Date(week),y=value), fill = "white", pch=21)+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")+xlab("")+geom_vline(xintercept = max(observed_data$date), col="grey", lty=2)

g1
```

```{r, fig.height=8, warning=FALSE, include=FALSE}
g.var/gH/g1
```
